---
title: "Evenness-Richness scatter plots: a graphical, intuitive approach to alpha diversity analysis"
subtitle: "Source code used for generating figures and tables"
author: "Jeff Gauthier, Nicolas Derome"
date: 'Last edit: September 2020'
output: pdf_document
---

# Introduction

```{r table1, tidy=TRUE}
# build a function to calculate Shannon's index:
get_shannon <- function(x) { # x is a numeric vector containing relative abundances
  h <- -sum(x*log(x))
  return(h)
}

# community number 1 (more even)
biome_1 <- data.frame(Species=c("sp1", "sp2", "sp3", "sp4", "sp5"),
                      prop=c(0.2, 0.2, 0.2, 0.2, 0.2))
biome_1
paste0("Shannon index: ", round(get_shannon(biome_1$prop), 3))

# community number 2 (more rich)
biome_2 <- data.frame(Species=c("sp1", "sp2", "sp3", "sp4", "sp5", "sp6", "sp7", "sp8"),
                      prop=c(0.5, 0.1, 0.1, 0.1, 0.095, 0.05, 0.03, 0.025))
biome_2
paste0("Shannon index: ", round(get_shannon(biome_2$prop), 3))
```


*********************************************

# Methodology

## Derivation of evenness

```{r table2, tidy=TRUE}
# community number 1 (more even)
biome_1$prop_ln_prop <- biome_1$prop * -log(biome_1$prop)
biome_1
paste0("Community 1 median evenness: ", median(biome_1$prop_ln_prop))
paste0("Community 1 normalized-median evenness: ", median(biome_1$prop_ln_prop) / max(biome_1$prop_ln_prop))

# community number 2 (more rich)
biome_2$prop_ln_prop <- biome_2$prop * -log(biome_2$prop)
biome_2
paste0("Community 2 median evenness: ", median(biome_2$prop_ln_prop))
paste0("Community 2 normalized-median evenness: ", median(biome_2$prop_ln_prop) / max(biome_2$prop_ln_prop))

# community number 3 (more even)
biome_3 <- data.frame(Species=c("sp1", "sp2", "sp3"),
                      prop=c(1/3, 1/3, 1/3))                    # ln_pi
biome_3$prop_ln_prop <- biome_3$prop * -log(biome_3$prop)      #pi ln pi
biome_3

# community number 4 (more rich)
biome_4 <- data.frame(Species=c("sp1", "sp2", "sp3"),
                      prop=c(0.95, 0.095, 0.005))             # ln_pi
biome_4$prop_ln_prop <- biome_4$prop * -log(biome_4$prop)    #pi ln pi
biome_4

# normalized-median evenness
norm_1 <- median(biome_1$prop_ln_prop) / max(biome_1$prop_ln_prop)
norm_2 <- median(biome_2$prop_ln_prop) / max(biome_2$prop_ln_prop)
norm_3 <- median(biome_3$prop_ln_prop) / max(biome_3$prop_ln_prop)
norm_4 <- median(biome_4$prop_ln_prop) / max(biome_4$prop_ln_prop)

# View in data frame
even_compare <- data.frame(Biome=c("Community 1", "Community 2", "Community 3", "Community 4"),
                           Richness=c(5, 8, 3, 3),
                           Norm_Even=c(norm_1, norm_2, norm_3, norm_4))
even_compare
```

## Graphical representation

```{r fig1, tidy=TRUE}
library(ggplot2)

fig1 <- ggplot(even_compare, aes(x=Richness, y=Norm_Even)) +
          geom_point() +
          geom_text(label=even_compare$Biome, nudge_y=-0.05) +
          xlim(0, 10) + ylim(0,1) +
          xlab("Species richness") + ylab("Normalized-mean evenness") +
          theme_bw()
fig1

# pdf
if(!dir.exists("Figures")) {dir.create("Figures")}
ggsave(fig1, filename="Figures/Figure_1.pdf")
```
 

## Example with a larger mock dataset

```{r table3, tidy=TRUE}
mydata <- read.table("01_fictional_data/fictional_data_v1.1.csv", header=TRUE)
mydata
```

```{r fig2, tidy=TRUE}
# make the 2D plot
library(ggplot2)

## Figure 2A
fig2a <- ggplot(mydata, aes(x=richness, y=norm_even, color=Group)) +
           geom_point() +
           geom_text(label=mydata$Biome) +
           xlab("Species richness") + ylab("Normalized-mean evenness") +
           stat_ellipse(level=0.95) +
           theme_bw() +
           ggtitle("A") +
           theme(legend.position="none")
fig2a

##Figure 2B
source("R_modules/summarySE.R")
fig2b.data <- summarySE(mydata, measurevar="shannon", groupvars=c("Group")) 
fig2b <- ggplot(fig2b.data, aes(x=Group, y=shannon, fill=Group)) +
        geom_bar(position=position_dodge(), stat="identity") +
        ylab("Shannon") +
        geom_errorbar(aes(ymin=shannon-ci, ymax=shannon+ci), width=.2) + 
        ggtitle("B")
fig2b

## grid arrange
library(gridExtra)
fig2 <- grid.arrange(fig2a, fig2b, ncol=2)
fig2

# pdf
if(!dir.exists("Figures")) {dir.create("Figures")}
ggsave(fig2, filename="Figures/Figure_2.pdf")
```

```{r permanova_mock_data, tidy=TRUE}
## add new grouping variable
mydata$Region <- rep(c("Urban", "Rural"), 5)

## table 3
mydata

## permanova
library(vegan)
set.seed(12345)
p3.permanova <- adonis(richness+norm_even~Group*Region, data=mydata, method="euclidean", permutations=99)
p3.permanova

## comparison with anova
p3.aov <- aov(shannon~Group*Region, data=mydata)
summary(p3.aov)
```


*************************************************

# Test on real-world data: Enterotypes of the human gut microbiome (2011)

## Introduction to the dataset

```{r ent_data_prep, tidy=TRUE}
library(phyloseq)
data(enterotype)

# Compile an ASV table with relative genera abundances per sample
ent.data <- otu_table(transform_sample_counts(enterotype, function(x) {x/sum(x)})) 
ent.data <- as.data.frame(ent.data)

# transpose otu_table (so that samples are rows)
ent.data <- t(ent.data)
ent.data <- as.data.frame(ent.data)

# replace otu names by unique number IDs
names(ent.data) <- seq(1:ncol(ent.data))

# product-ln-transform sample counts
plp.ent.data <- - ent.data * log(ent.data)

# replace NaN by zero
plp.ent.data[plp.ent.data=="NaN"] <- 0

# add Evenness (median must not include zero elements!)
library(matrixStats)
library(qlcMatrix)
plp.ent.data$Evenness <- apply(plp.ent.data, 1, function(x) median(x[x != 0])/max(x))

# add Shannon index
plp.ent.data$Shannon <- rowSums(plp.ent.data[, 1:553])

# add observed richness (only rows 1:54 contain counts!)
# rowSums(c!=0) counts non-zero cells pre each row
plp.ent.data$Richness <- rowSums(plp.ent.data[, 1:553] != 0)

# Add sequencing tech
plp.ent.data$SeqTech <- sample_data(enterotype)$SeqTech[1:280]

# OTU abundances are nor required anymore
plp.ent.data[, 1:553] <- NULL

# view table
plp.ent.data
```

## Basic alpha-diversity analysis

```{r fig3, tidy=TRUE}

# figure 3a
# Add sample group variable
plp.ent.data$Enterotype <- sample_data(enterotype)$Enterotype[1:280]
# use NA-free dataset (UC and CD removed because n=1)
plp.ent.data.noNA <- subset(plp.ent.data, plp.ent.data$Enterotype %in% c("1", "2", "3"))
# top 10 taxa
source("R_modules/plot_bar_err.R")
fig3a <- plot_bar_err(subset_samples(enterotype, Enterotype %in% c("1", "2", "3")), 
                      by="Enterotype", n=10, top.type="taxa", taxrank="Genus", facet=~Enterotype, ncol=1) +
               ylab("Relative abundance") +
               ggtitle("A")
fig3a

# figure 3b
source("R_modules/summarySE.R")
# summarize data and remove NA's
plp.ent.data.noNA.summary <- summarySE(plp.ent.data.noNA, measurevar="Shannon", groupvars=c("Enterotype"))
# make the shannon index plot
fig3b <- ggplot(plp.ent.data.noNA.summary, aes(x=Enterotype, y=Shannon, fill=Enterotype)) +
        geom_bar(position=position_dodge(), stat="identity") +
        geom_errorbar(aes(ymin=Shannon-ci, ymax=Shannon+ci), width=.2) +
        ggtitle("B")
fig3b

#figure 3c
# make evenness vs richness plot
library(ggplot2)
fig3c <- ggplot(plp.ent.data.noNA, aes(x=Richness, y=Evenness, color=Enterotype)) +
                       geom_point() +
                       stat_ellipse(level=0.95) +
                       theme_bw() +
                       ggtitle("C")
fig3c

# grid arrange
library(gridExtra)
fig3 <- grid.arrange(fig3a, grid.arrange(fig3b, fig3c, ncol=1), ncol=2)

# pdf
if(!dir.exists("Figures")) {dir.create("Figures")}
ggsave(fig3, filename="Figures/Figure_3.pdf")
```

```{r chi-squared_test_on_clusters, tidy=TRUE}
## add EvR clusters info to dataset
plp.ent.data.noNA$Cluster <- ifelse(plp.ent.data.noNA$Richness > 400, "Cluster 2", "Cluster 1")

## summarize data by cluster
library(reshape2)
cluster.data <- dcast(plp.ent.data.noNA, formula = Cluster~Enterotype)
colnames(cluster.data) <- c("Cluster", "Samples from Enterotype 1", "Samples from Enterotype 2", "Samples from Enterotype 3")

## add margin totals
cluster.data$Total <- cluster.data$`Samples from Enterotype 1` + 
                      cluster.data$`Samples from Enterotype 2` + 
                      cluster.data$`Samples from Enterotype 3`

cluster.data[3,] <- c("Total", sum(cluster.data$`Samples from Enterotype 1`), 
                               sum(cluster.data$`Samples from Enterotype 2`), 
                               sum(cluster.data$`Samples from Enterotype 3`), 
                               sum(cluster.data$Total))

## View table 4
cluster.data

## Independence between 2 clusters (chi-squared)
chisqdf <- cluster.data[1:2, 2:4]
rownames(chisqdf) <- c("Cluster 1", "Cluster 2")
X <- chisq.test(data.matrix(chisqdf))
```

## Clustering effect caused by sequencing technology

```{r fig4, tidy=TRUE}
# figure 4a
library(ggplot2)
fig4a <- ggplot(plp.ent.data.noNA, aes(x=Richness, y=Evenness, color=SeqTech)) +
                       geom_point() +
                       stat_ellipse(level=0.95) +
                       theme_bw() +
                       ggtitle("A") +
                       theme(legend.position = "none")
fig4a

# figure 4b
source("R_modules/summarySE.R")
# make summary
plp.ent.data.noNA.summary <- summarySE(plp.ent.data.noNA, measurevar="Shannon", groupvars=c("SeqTech")) 
# make shannon index plot
fig4b <- ggplot(plp.ent.data.noNA.summary, aes(x=SeqTech, y=Shannon, fill=SeqTech)) +
        geom_bar(position=position_dodge(), stat="identity") +
        geom_errorbar(aes(ymin=Shannon-ci, ymax=Shannon+ci), width=.2) +
        ggtitle("B")
fig4b

# grid arrange
library(gridExtra)
fig4 <- grid.arrange(fig4a, fig4b, ncol=2)

# pdf
if(!dir.exists("Figures")) {dir.create("Figures")}
ggsave(fig4, filename="Figures/Figure_4.pdf")
```

```{r fig5, tidy=TRUE}
# make the 2D plot
library(ggplot2)

# cluster1
fig5a <- ggplot(subset(plp.ent.data.noNA, plp.ent.data.noNA$Cluster %in% "Cluster 1"), aes(x=Richness, y=Evenness, color=Enterotype)) +
                       geom_point() +
                       stat_ellipse(level=0.95) +
                       ggtitle("Cluster 1 (Sanger/454)") +
                       theme_bw()

# cluster2
fig5b <- ggplot(subset(plp.ent.data.noNA, plp.ent.data.noNA$Cluster %in% "Cluster 2"), aes(x=Richness, y=Evenness, color=Enterotype)) +
                       geom_point() +
                       stat_ellipse(level=0.95) +
                       ggtitle("Cluster 2 (Illumina)") +
                       theme_bw()

# show plot
library(gridExtra)
fig5 <- grid.arrange(fig5a, fig5b)
fig5

# pdf
if(!dir.exists("Figures")) {dir.create("Figures")}
ggsave(fig5, filename="Figures/Figure_5.pdf")
```

## Use of PERMANOVA to verify the clustering effect

```{r permanova_enterotype, tidy=TRUE}
library(vegan)
set.seed(12345)
ent.permanova <- adonis(Evenness+Richness~Cluster*Enterotype, data=plp.ent.data.noNA, method = "euclidean", permutations = 99)
ent.permanova
```

## Influence of sequencing technology on top genera

```{r fig6, tidy=TRUE}
source("R_modules/plot_bar_err.R")

# custom color palette
color_bacteroides <- "red3"
color_bifidobacterium <- "forestgreen"
color_faecalibacterium <- "turquoise2"
color_lachnospira <- "blue3"
color_prevotella <- "magenta"
color_blautia <- "slategray"
color_roseburia <- "gold4"

# split data by clusters: 1 (Sanger/454) and 2 (Illumina)
enterotype.sanger <- subset_samples(enterotype, SeqTech %in% "Sanger")
enterotype.454 <- subset_samples(enterotype, SeqTech %in% "Pyro454")
enterotype.illumina <- subset_samples(enterotype, SeqTech %in% "Illumina")

# top 10 taxa (Sanger)
fig6.sanger <- plot_bar_err(subset_samples(enterotype.sanger, Enterotype %in% c("1", "2", "3")), 
                             by="Enterotype", n=5, top.type="taxa", taxrank="Genus", facet=~Enterotype, ncol=1) +
  ggtitle("Sanger") +
  theme(legend.position="bottom", legend.direction="vertical") +
  scale_y_continuous(breaks=c(0, 0.5, 1), limits=c(0,1), name="Relative abundance") +
  scale_fill_manual(values=c(color_bacteroides,
                              color_bifidobacterium,
                              color_faecalibacterium,
                              color_lachnospira,
                              color_prevotella))

# top 10 taxa (454)
fig6.454 <- plot_bar_err(subset_samples(enterotype.454, Enterotype %in% c("1", "2", "3")), 
                          by="Enterotype", n=5, top.type="taxa", taxrank="Genus", facet=~Enterotype, ncol=1) +
  ggtitle("454") +
  theme(legend.position="bottom", legend.direction="vertical") +
  scale_y_continuous(breaks=c(0, 0.5, 1), limits=c(0,1), name=NULL) +
  scale_fill_manual(values=c(color_bacteroides,
                              color_blautia,
                              color_faecalibacterium,
                              color_prevotella,
                              color_roseburia))

# top 10 taxa (Illumina)
fig6.illumina <- plot_bar_err(subset_samples(enterotype.illumina, Enterotype %in% c("1", "2", "3")), 
                               by="Enterotype", n=5, top.type="taxa", taxrank="Genus", facet=~Enterotype, ncol=1) +
  ggtitle("Illumina") +
  theme(legend.position="bottom", legend.direction="vertical") +
  scale_y_continuous(breaks=c(0, 0.5, 1), limits=c(0,1), name=NULL) +
  scale_fill_manual(values=c(color_bacteroides,
                              color_faecalibacterium,
                              color_lachnospira,
                              color_prevotella,
                              color_roseburia))

# arrange and view plot
library(gridExtra)
fig6 <- grid.arrange(arrangeGrob(fig6.sanger, fig6.454, ncol=2), 
             arrangeGrob(fig6.illumina, nrow=1), nrow=1, widths=c(2,1))
fig6

# pdf
if(!dir.exists("Figures")) {dir.create("Figures")}
ggsave(fig6, filename="Figures/Figure_6.pdf")
```
